Boundary terms in the action for Causal Sets

### 2014-09-18 test6.tex
initial commit

Given a causet, $\mathcal{C}$, with a spacelike partition so that $\mathcal C = \mathcal C^+ \cup\, \mathcal C^-$, we propose the following family of causal set operators for calculating the GHY boundary term of this partition:
\be\label{general_boundary_sum}
\mathcal{S}^{(d)}_{GHY}\left[\mathcal{C}^-,\mathcal{C}^+;\mathbf{p}, \mathbf{q} \right]= \left(l/l_p\right)^{d-2} c_{d}
\left( \sum_m p_m N_{max}^{(m)}\left[\mathcal{C}^- \right]
+  \sum_n q_n N_{min}^{(n)}\left[\mathcal{C}^+ \right]\right)
\ee
where the constant $c_{d}$ only depends on the spacetime dimension and is given by
\be\label{Cn}
c_{d}=\frac{d(d+1)}{(d+2)}\left[\frac{V_{d-1}}{d}\right]^{\frac{2}{d}},
\ee
and where $V_d=\pi^{\frac{d}{2}}/\Gamma\left(\frac{d}{2}+1\right)$ denotes the volume of the unit $d$-ball ($\Gamma(t)$ is a gamma function, defined as $\Gamma(t):=\int_0^\infty dx\: x^{t-1}\: e^{-x}$). We have used $\mathbf{p}$ and $\mathbf{q}$ to stand for infinite dimensional vectors $(p_0,...,p_m,..)$ and $(q_0,...,q_n,..)$ respectively, where all the $p_m,\: q_n \in \mathbb{R}$. These vectors contain a finite number of non-zero entries, corresponding to the coefficients of the different $N_{{max\vphantom{i}}}^{\:(m)}$ and $N_{{min}}^{\:(n)}$ terms. Their values can be chosen arbitrarily, so long as they satisfy the following conditions.
\begin{align}\label{coefficient_relation1}
& \sum_m p_m \frac{\Gamma\left(\frac{1}{d}+m \right)}{m!}  + \sum_n q_n\frac{\Gamma\left(\frac{1}{d}+n \right)}{n!}=0
\\
& \label{coefficient_relation2}\sum_m p_m \frac{\Gamma\left(\frac{2}{d}+m \right)}{m!}  - \sum_n q_n\frac{\Gamma\left(\frac{2}{d}+n \right)}{n!}=1
\end{align}
The aforementioned family simply corresponds to all the different vectors, $\mathbf{p}$ and $\mathbf{q}$, that satisfy the above criteria.

@@

Given our ignorance in this matter,  we can ask if causal set theory  might  in fact guide us in the right direction.   Our starting point is the $d$-dimensional  Benincasa-Dowker Action \cite{Benincasa_Dowker:The_Scalar_Curvature_of_a_Causal_Set, Dowker_Glaser:dAlembertians_for_Causal_Sets} which is constructed from the mean number  $N_i^{(d)}${\footnote{ $N_i^{(d)}$ is dimension  and geometry dependent, but for our purpose it suffices to retain the dimension label. } of $i+1$-element inclusive intervals in the causal set 
\begin{equation}   
8 \pi G {S_{BD}^{(d)}}[\mathcal C] = -\alpha_d \rho^{\frac{2-d}{d}}\biggl( N[\mathcal C]+ \frac{\beta_d}{ \alpha_d} \sum_{i=1}^{n_d-1}  C^{(d)}_{i} N _i^{(d)}[\mathcal C] \biggr) 
\label{bd} 
\end{equation}  
with   
\begin{equation}
\begin{aligned}
 \alpha_d =   
\begin{cases}
-\frac{1}{\Gamma(\frac{2+d}{d})}c_d ^{2/d}  \quad   &d \, \, \mathrm{odd} \\
-2 \frac{1}{\Gamma(\frac{2+d}{d})}c_d ^{2/d}  \quad   &d \, \,  \mathrm{even} \\
\end{cases}
\qquad\text{and}\qquad \beta_d = 
\begin{cases} 
\frac{d+1}{2^{d-1}\Gamma(2/d +1)} c_d^{2/d}   \quad   &d\mathrm{ \, \, odd}\\ 
\frac{\Gamma(d/2+2)\Gamma(d/2+1)}{\Gamma(2/d)\Gamma(d)}c_d^{2/d}  &  d\mathrm{ \, \,  even}\\ 
\end{cases} 
\end{aligned}
\end{equation}
and
\begin{equation} 
n_d = 
\begin{cases} 
\frac{d-1}{2} + 2  \quad & d\mathrm{ \, \, odd}\\ 
\frac{d}{2} + 2  \quad &d\mathrm{ \, \,  even,}\\ 
\end{cases} 
\end{equation} 
where  $2^{-d/2} c_d$ is the volume of a unit Alexandrov interval $I(x,y)$, i.e., one for which the proper time $\tau(x,y)=1$. 
%%{ {\bf \color{red} this is not correct! $c_d= \vol(S^{d-2}) \frac{ 2^{\frac{2-d}{2}}}{d(d-1)}$ and $\vol(S^{d-2})$ the volume of the homogeneous $d-2$ sphere}}.  
The coefficients $C_i^{(d)}$ of the terms $N_i^{(d)}$ in the sum are
\begin{equation}
\label{cid}
 C_i^{(d)}= 
\begin{cases} 
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma(d/2(k+1)+3/2)}{\Gamma(d/2+3/2) \Gamma(1+dk/2)}  \quad  &\mathrm{d \, \, odd}\\ 
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma(d/2(k+1)+2)}{\Gamma(d/2+2) \Gamma(1+dk/2)}  & \mathrm{d \, \,  even.}\\ 
\end{cases} 
\end{equation} 
These coefficients can be expressed more compactly as hypergeometric functions 
\begin{equation}
C_i^d={}_{q+1}F_{q}(\{ a_1, \ldots a_q, i-1\}, \{b_1, \ldots b_q \} |1)
\label{chyp} 
\end{equation} 
with $q=d/2+1/2, a_i=\frac{d+2i}{d}, b_i=2i/d$ for $d$ odd and $q=d/2, a_i=\frac{d+2i+2}{d}, b_i=2i/d$  for $d$ even.

@@

In \cite{Glaser_Sumati:Locality_in_Causal_Set} a closed form expression was obtained for the mean value of the number of $i+1$ element inclusive intervals $N_i^d$ contained in an Alexandrov interval in $d$ dimensional flat spacetime 
\begin{equation} N_i^d= \frac{\Gamma(d)^2\, \, N^{i+2}}{\Gamma(i)} \sum_{k=0}^\infty \frac{(-N)^k \, \, \, \Gamma(\frac{d(k+i)}{2}+1) \, \, \,  \Gamma(\frac{d(k+i+1)}{2}+1)}{(k+i+2) \, (k+i+1)\, \Gamma(k+1)\, \Gamma(\frac{d(k+i)}{2} +d) \,\Gamma(\frac{d(k+i+1)}{2}+d)}, 
\end{equation} 
where $i \geq 1$.  We use this to expand the sum in (\ref{bd}) as a power series in $N$: $\sum_{i=1}^{n_d} C_i^d \, \, N_i^{(d)} = \sum_{j=1}^\infty A_j^d \, \, N^{j+1}$.  After a rearrangement and redefinition of indices we find that 
\begin{equation}
 A_j^d = \Gamma(d)^2\biggl( \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d  \biggr) \frac{(-1)^j}{(j+1)!}\frac{\Gamma(\frac{d}{2}(j-1)+1)\Gamma(\frac{d}{2}j+1)}{\Gamma(\frac{d}{2}(j-1)+d) \Gamma(\frac{d}{2}j+d)}, 
\end{equation} 
where $d=2D$ for $d$ even and $d=2D+1$ for $d$ odd.   
% Using the abbreviated form $C_i^d=\sum_{l=0}^{i-1} \binom{i-1}{l}A(d,l)$ where $A(d,l)$ is defined using \ref{cid} 
% we find that 
% \begin{equation}
% \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d = (-1)^{j-1} A(d,j-1),  
% \end{equation}      
% where we have used the binomial identities  BLAH. Thus 
% \begin{equation}
%  \sum_{i=1}^{n_d} C_i^d N_i^{(d)}= ((d-1)!)^2 \sum_{j=1}^\infty(-1)^j B(d,j) \frac{\Gamma(\frac{d}{2}j+1)}{\Gamma(j+2)\Gamma(\frac{d}{2}(j-1)+d)\Gamma(\frac{d}{2}j+d)}N^{j+1}
% \label{thesum} 
% \end{equation} 
% where 
% \begin{equation} 
% B(d,j) = \left \{ 
% \begin{array}{ccc} 
% \frac{\Gamma(\frac{d}{2}j+3/2)}{\Gamma(\frac{d}{2}+3/2)}  &\quad & d \, \, odd, \\  \frac{\Gamma(\frac{d}{2}j+2)}{\Gamma(\frac{d}{2}+2)}  &\quad & d \, \, even.  \\ 
% \end{array} 
% \right
% \end{equation} 
This can be simplified to 
\begin{equation}
A_j^d= \gamma_j^d \, \, \Gamma(d)^2 (-1)^{j+1} \frac{1}{\Gamma(\frac{d}{2} (j + 1)) \Gamma(\frac{d}{2} (2 + j)) \Gamma(2 + j) }
\label{simplercoefft} 
\end{equation} 
where 
\begin{equation} \gamma_j^d= 
\begin{cases} 
\frac{\sqrt{\pi}}{2^{1+dj} }\frac{\Gamma(2+dj)}{\Gamma(\frac{d-1}{2})}    \quad   \mathrm{d} \, \, \mathrm{odd} \\
\frac{\Gamma(1+\frac{d}{2}j)  \Gamma(2+\frac{d}{2}j) }{\Gamma(\frac{d}{2})}  \quad   \mathrm{d \, \,  even} \\
\end{cases} 
\end{equation} 
A further simplification then allows us to include the first term in the action (\ref{bd}) in the series expansion so that 
\begin{equation} 
8 \pi G S_{BD}^d(C) = -\alpha_d \rho^{\frac{2-d}{d}} \sum_{j=0}^\infty (-1)^j Q_j^d N^{j+1} 
\end{equation}
where 
\begin{eqnarray} 
Q_j^d & =& {\widetilde \gamma}_j^d \, \, \frac{\Gamma(d)\Gamma(\frac{d}{2})}{\Gamma(\frac{d}{2}(1+j)) \Gamma(\frac{d}{2}(2+j)) \Gamma(2+j)} \\ \nonumber 
{\widetilde \gamma}_j^d &=& 
\begin{cases} 
2^{-dj}    \quad   \mathrm{d} \, \, \mathrm{odd} \\
\Gamma(1+\frac{d}{2}j) \Gamma(2+\frac{d}{2}j)   \quad   \mathrm{d \, \,  even} \\
\end{cases} 
\label{fullcoefft} 
\end{eqnarray} 
Taking our cues from the behaviour of $8 \pi G S^d(C)$ for $d=2,3$ in the  $N \rightarrow \infty$ limit,  we will consider the ratio
\begin{equation} 
\frac{ 8 \pi G S^d(C)}{\vol(S^{d-2})} = \frac{b}{d(d-1) \Gamma(1+2/d) N^{\frac{d-2}{d}}} \left( N + \frac{\beta_d}{\alpha_d} \sum_{i=1}^{n_d} C_i^d N_i^d \right)  
\label{ratio} 
\end{equation} 
where $b=1$ for $d$ odd and $b=2$ for $d$ even.
Using this simplification an explicit calculation in Mathematica\footnote{The above simplifications greatly aid the Mathematica calculation and allow us to extend it to higher $d$.}  for $d=2, \ldots 16$ gives 
\begin{equation} 
\lim_{N \rightarrow \infty} \frac{ 8 \pi G S_{BD}^d(C)}{\vol(S^{d-2})} = 1,  
\label{result} 
\end{equation}

>>>>

Given a causet, $\mathcal{C}$, with a spacelike partition so that $\mathcal C = \mathcal C^+ \cup\, \mathcal C^-$, we propose the following family of causal set operators for calculating the GHY boundary term of this partition:
\be\label{general_boundary_sum}
\mathcal{S}^{(d)}_{GHY}\left[\mathcal{C}^-,\mathcal{C}^+;\mathbf{p}, \mathbf{q} \right]= \left(l/l_p\right)^{d-2} c_{d}
\left( \sum_m p_m N_{max}^{(m)}\left[\mathcal{C}^- \right]
+  \sum_n q_n N_{min}^{(n)}\left[\mathcal{C}^+ \right]\right)
\ee
where the constant $c_{d}$ only depends on the spacetime dimension and is given by
\be\label{Cn}
c_{d}=\frac{d(d+1)}{(d+2)}\left[\frac{V_{d-1}}{d}\right]^{\frac{2}{d}},
\ee
and where $V_d=\pi^{\frac{d}{2}}/\Gamma\left(\frac{d}{2}+1\right)$ denotes the volume of the unit $d$-ball ($\Gamma\left(t)$ is a gamma function, defined as $\Gamma(t):=\int_0^\infty dx\: x^{t-1}\: e^{-x}$). We have used $\mathbf{p}$ and $\mathbf{q}$ to stand for infinite dimensional vectors $(p_0,...,p_m,..)$ and $(q_0,...,q_n,..)$ respectively, where all the $p_m,\: q_n \in \mathbb{R}$. These vectors contain a finite number of non-zero entries, corresponding to the coefficients of the different $N_{{max\vphantom{i}}}^{\:(m)}$ and $N_{{min}}^{\:(n\right)}$ terms. Their values can be chosen arbitrarily, so long as they satisfy the following conditions.
\begin{align}\label{coefficient_relation1}
& \sum_m p_m \frac{\Gamma\left(\frac{1}{d}+m \right)}{m!}  + \sum_n q_n\frac{\Gamma\left(\frac{1}{d}+n \right)}{n!}=0
\\
& \label{coefficient_relation2}\sum_m p_m \frac{\Gamma\left(\frac{2}{d}+m \right)}{m!}  - \sum_n q_n\frac{\Gamma\left(\frac{2}{d}+n \right)}{n!}=1
\end{align}
The aforementioned family simply corresponds to all the different vectors, $\mathbf{p}$ and $\mathbf{q}$, that satisfy the above criteria.

@@

Given our ignorance in this matter,  we can ask if causal set theory  might  in fact guide us in the right direction.   Our starting point is the $d$-dimensional  Benincasa-Dowker Action \cite{Benincasa_Dowker:The_Scalar_Curvature_of_a_Causal_Set, Dowker_Glaser:dAlembertians_for_Causal_Sets} which is constructed from the mean number  $N_i^{(d)}${\footnote{ $N_i^{(d)}$ is dimension  and geometry dependent, but for our purpose it suffices to retain the dimension label. } of $i+1$-element inclusive intervals in the causal set 
\begin{equation}   
8 \pi G {S_{BD}^{(d)}}[\mathcal C] = -\alpha_d \rho^{\frac{2-d}{d}}\biggl( N[\mathcal C]+ \frac{\beta_d}{ \alpha_d} \sum_{i=1}^{n_d-1}  C^{(d)}_{i} N _i^{(d)}[\mathcal C] \biggr) 
\label{bd} 
\end{equation}  
with   
\begin{equation}
\begin{aligned}
 \alpha_d =   
\begin{cases}
-\frac{1}{\Gamma\left(\frac{2+d}{d}\right)}c_d ^{2/d}  \quad   &d \, \, \mathrm{odd} \\
-2 \frac{1}{\Gamma\left(\frac{2+d}{d}\right)}c_d ^{2/d}  \quad   &d \, \,  \mathrm{even} \\
\end{cases}
\qquad\text{and}\qquad \beta_d = 
\begin{cases} 
\frac{d+1}{2^{d-1}\Gamma\left(2/d +1\right)} c_d^{2/d}   \quad   &d\mathrm{ \, \, odd}\\ 
\frac{\Gamma\left(d/2+2)\Gamma(d/2+1)}{\Gamma(2/d)\Gamma(d\right)}c_d^{2/d}  &  d\mathrm{ \, \,  even}\\ 
\end{cases} 
\end{aligned}
\end{equation}
and
\begin{equation} 
n_d = 
\begin{cases} 
\frac{d-1}{2} + 2  \quad & d\mathrm{ \, \, odd}\\ 
\frac{d}{2} + 2  \quad &d\mathrm{ \, \,  even,}\\ 
\end{cases} 
\end{equation} 
where  $2^{-d/2} c_d$ is the volume of a unit Alexandrov interval $I(x,y)$, i.e., one for which the proper time $\tau(x,y)=1$. 
%%{ {\bf \color{red} this is not correct! $c_d= \vol(S^{d-2}) \frac{ 2^{\frac{2-d}{2}}}{d(d-1)}$ and $\vol(S^{d-2})$ the volume of the homogeneous $d-2$ sphere}}.  
The coefficients $C_i^{(d)}$ of the terms $N_i^{(d)}$ in the sum are
\begin{equation}
\label{cid}
 C_i^{(d)}= 
\begin{cases} 
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma\left(d/2(k+1)+3/2)}{\Gamma(d/2+3/2) \Gamma(1+dk/2\right)}  \quad  &\mathrm{d \, \, odd}\\ 
\sum_{k=0}^{i-1} (-1)^k\binom{i-1}{k} \frac{\Gamma\left(d/2(k+1)+2)}{\Gamma(d/2+2) \Gamma(1+dk/2\right)}  & \mathrm{d \, \,  even.}\\ 
\end{cases} 
\end{equation} 
These coefficients can be expressed more compactly as hypergeometric functions 
\begin{equation}
C_i^d={}_{q+1}F_{q}(\{ a_1, \ldots a_q, i-1\}, \{b_1, \ldots b_q \} |1)
\label{chyp} 
\end{equation} 
with $q=d/2+1/2, a_i=\frac{d+2i}{d}, b_i=2i/d$ for $d$ odd and $q=d/2, a_i=\frac{d+2i+2}{d}, b_i=2i/d$  for $d$ even.

@@

In \cite{Glaser_Sumati:Locality_in_Causal_Set} a closed form expression was obtained for the mean value of the number of $i+1$ element inclusive intervals $N_i^d$ contained in an Alexandrov interval in $d$ dimensional flat spacetime 
\begin{equation} N_i^d= \frac{\Gamma\left(d)^2\, \, N^{i+2}}{\Gamma(i)} \sum_{k=0}^\infty \frac{(-N)^k \, \, \, \Gamma(\frac{d(k+i)}{2}+1) \, \, \,  \Gamma(\frac{d(k+i+1)}{2}+1)}{(k+i+2) \, (k+i+1)\, \Gamma(k+1)\, \Gamma(\frac{d(k+i)}{2} +d) \,\Gamma(\frac{d(k+i+1)}{2}+d\right)}, 
\end{equation} 
where $i \geq 1$.  We use this to expand the sum in (\ref{bd}) as a power series in $N$: $\sum_{i=1}^{n_d} C_i^d \, \, N_i^{(d)} = \sum_{j=1}^\infty A_j^d \, \, N^{j+1}$.  After a rearrangement and redefinition of indices we find that 
\begin{equation}
 A_j^d = \Gamma\left(d)^2\biggl( \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d  \biggr) \frac{(-1)^j}{(j+1)!}\frac{\Gamma(\frac{d}{2}(j-1)+1)\Gamma(\frac{d}{2}j+1)}{\Gamma(\frac{d}{2}(j-1)+d) \Gamma(\frac{d}{2}j+d\right)}, 
\end{equation} 
where $d=2D$ for $d$ even and $d=2D+1$ for $d$ odd.   
% Using the abbreviated form $C_i^d=\sum_{l=0}^{i-1} \binom{i-1}{l}A(d,l)$ where $A(d,l)$ is defined using \ref{cid} 
% we find that 
% \begin{equation}
% \sum_{i=1}^{D+2} (-1)^i \binom{j-1}{i-1}  C_i^d = (-1)^{j-1} A(d,j-1),  
% \end{equation}      
% where we have used the binomial identities  BLAH. Thus 
% \begin{equation}
%  \sum_{i=1}^{n_d} C_i^d N_i^{(d)}= ((d-1)!)^2 \sum_{j=1}^\infty(-1)^j B(d,j) \frac{\Gamma\left(\frac{d}{2}j+1)}{\Gamma(j+2)\Gamma(\frac{d}{2}(j-1)+d)\Gamma(\frac{d}{2}j+d\right)}N^{j+1}
% \label{thesum} 
% \end{equation} 
% where 
% \begin{equation} 
% B(d,j) = \left \{ 
% \begin{array}{ccc} 
% \frac{\Gamma\left(\frac{d}{2}j+3/2)}{\Gamma(\frac{d}{2}+3/2)}  &\quad & d \, \, odd, \\  \frac{\Gamma(\frac{d}{2}j+2)}{\Gamma(\frac{d}{2}+2\right)}  &\quad & d \, \, even.  \\ 
% \end{array} 
% \right
% \end{equation} 
This can be simplified to 
\begin{equation}
A_j^d= \gamma_j^d \, \, \Gamma\left(d)^2 (-1)^{j+1} \frac{1}{\Gamma(\frac{d}{2} (j + 1)) \Gamma(\frac{d}{2} (2 + j)) \Gamma(2 + j\right) }
\label{simplercoefft} 
\end{equation} 
where 
\begin{equation} \gamma_j^d= 
\begin{cases} 
\frac{\sqrt{\pi}}{2^{1+dj} }\frac{\Gamma\left(2+dj)}{\Gamma(\frac{d-1}{2}\right)}    \quad   \mathrm{d} \, \, \mathrm{odd} \\
\frac{\Gamma\left(1+\frac{d}{2}j)  \Gamma(2+\frac{d}{2}j) }{\Gamma(\frac{d}{2}\right)}  \quad   \mathrm{d \, \,  even} \\
\end{cases} 
\end{equation} 
A further simplification then allows us to include the first term in the action (\ref{bd}) in the series expansion so that 
\begin{equation} 
8 \pi G S_{BD}^d(C) = -\alpha_d \rho^{\frac{2-d}{d}} \sum_{j=0}^\infty (-1)^j Q_j^d N^{j+1} 
\end{equation}
where 
\begin{eqnarray} 
Q_j^d & =& {\widetilde \gamma}_j^d \, \, \frac{\Gamma\left(d)\Gamma(\frac{d}{2})}{\Gamma(\frac{d}{2}(1+j)) \Gamma(\frac{d}{2}(2+j)) \Gamma(2+j\right)} \\ \nonumber 
{\widetilde \gamma}_j^d &=& 
\begin{cases} 
2^{-dj}    \quad   \mathrm{d} \, \, \mathrm{odd} \\
\Gamma\left(1+\frac{d}{2}j) \Gamma(2+\frac{d}{2}j\right)   \quad   \mathrm{d \, \,  even} \\
\end{cases} 
\label{fullcoefft} 
\end{eqnarray} 
Taking our cues from the behaviour of $8 \pi G S^d(C)$ for $d=2,3$ in the  $N \rightarrow \infty$ limit,  we will consider the ratio
\begin{equation} 
\frac{ 8 \pi G S^d(C)}{\vol(S^{d-2})} = \frac{b}{d(d-1) \Gamma\left(1+2/d) N^{\frac{d-2}{d}}} \left( N + \frac{\beta_d}{\alpha_d} \sum_{i=1}^{n_d} C_i^d N_i^d \right\right)  
\label{ratio} 
\end{equation} 
where $b=1$ for $d$ odd and $b=2$ for $d$ even.
Using this simplification an explicit calculation in Mathematica\footnote{The above simplifications greatly aid the Mathematica calculation and allow us to extend it to higher $d$.}  for $d=2, \ldots 16$ gives 
\begin{equation} 
\lim_{N \rightarrow \infty} \frac{ 8 \pi G S_{BD}^d(C)}{\vol(S^{d-2})} = 1,  
\label{result} 
\end{equation}
